---
title: "Spatial Data Tutorial"
project:
  type: default
  output-dir: docs
format:
  gfm:
    toc: true
date: 2026-03-12
author: Allie Warren, allison.warren
editor: visual
prefer-html: true
---

## Overview

This tutorial walks through a real-world spatial data workflow using school and school district vaccination data from Washington State. The vaccination data was downloaded from the [WA DOH School Immunization Data Dashboard](https://doh.wa.gov/data-and-statistical-reports/washington-tracking-network-wtn/school-immunization/dashboard)and the location data is pulled from the [Washington State Geospatial Open Data Portal](https://geo.wa.gov/search?q=school). This tutorial shows how to:

-   Read Excel and spatial shape data files
-   Geocode addresses to latitude/longitude coordinates
-   Pull up-to-date spatial data using an API
-   Combine data sources
-   Create maps using `sf` and `ggplot2`

```{r}
# Package descriptions;
# readr: read in files
# dplyr, tidyr, stringr: data manipulation
# readxl: reading Excel files
# tidygeocoder: geocoding addresses
# ggplot2: plotting
# usmap: us mapping data
# sf: working with spatial/shape data
# viridis: color palettes
# httr2: making API requests
# jsonlite: parsing JSON responses
# janitor: data cleaning utilities
# xml2: Parse XML
# glue: used for creating strings with variables included

# use pacman for loading/installing the other packages
if(!require("pacman")) install.packages("pacman") 

# This load packages
pacman::p_load(readr, dplyr, tidyr, readxl, tidygeocoder, sf, httr2, stringr, ggplot2, usmap, sf, viridis, jsonlite, janitor, xml2, glue)
```

## Reading and Cleaning Data

Read excel can read xls or xlsx files. Especially with excel files, it is useful to run clean_names(), as excel column names can often have inconsistent formats. clean_names() adjusts the column names so they are unique and consist only of the \_ character, numbers, and letters.

### What is a Shapefile?

Spatial data describing geographic boundaries (like school districts) is often stored as a **shapefile** — a collection of files that together define the geometry (shapes) and attributes (data) of geographic features and can be used for creating map plots. Each feature in a shapefile has:

-   A **geometry**: the shape itself, stored as points, lines, or polygons
-   **Attributes**: a table of data associated with each shape (like a name or ID)

In R, the `sf` package represents this as a dataframe with a special `geometry` column.

```{r}
# vaccination data source: https://doh.wa.gov/data-and-statistical-reports/washington-tracking-network-wtn/school-immunization/dashboard

# read in school level vaccination data
school_data <- read_excel("../data/2024-2025SchoolBuilding.xlsx") |> 
  clean_names()

# read in school district level vaccination data
# the excel file includes state, county, and school district vaccination data, so we specify which sheet we want to read in
district_data <- read_excel("../data/2024-2025SchoolYear.xlsx", sheet = 'School District') |> 
  clean_names()

# read in shape data with the location of WA state public schools
# data source: https://geo.wa.gov/datasets/2b1977231086490a933ee8522dc85a76_0/explore?location=47.264054%2C-120.796522%2C7
school_location <- read_sf("../data/WashingtonStatePublicSchools/WashingtonStatePublicSchools.shp")
  
```

### Select Data

Select the vaccination data (condition, grade level, and vaccination status) we want to visualize.

```{r}
# select data for plotting:
# decide on condition (data contains Overall, Diptheria, Tetanus, Pertussis, Measles, Mumps, Rubella, Hepatitis B, Varicella, and Polion)
# and grade level (data contains Seventh Grade, K-12, and Kindergarten)
# and vaccination level (either Complete, Conditional, Out of Compliance, Exempt, Personal Exemption, Medical Exemption, Religious Exemption, or Religious Membership Excemption)
condition <- 'Measles'
select_grade <- 'K-12'
vaccination_level <- 'Complete'

# filter school level data to selected condition, grade level, and vaccination status 
school_data_condition <- filter(school_data, disease_or_vaccine == condition,
                                immunization_status == vaccination_level,
                                grade == select_grade)

# check that there is only one data point per school
if(nrow(school_data_condition) != nrow(school_data_condition |> select(school_name, school_district, address) |> distinct())) {
  print('Multiple data points per school, filtering to just one record per school')
  # and view any duplicates
 print(school_data_condition |>
    filter(duplicated(cbind(school_name, school_district, address)) | duplicated(cbind(school_name, school_district, address), fromLast = TRUE)) |> 
   select(school_name, count, enrollment, school_district, address))
  
  # remove duplicates, keeping only the school with the higher enrollment
  school_data_condition <- school_data_condition |> 
    slice_max(enrollment, by = c(school_name, school_district, address), n = 1, with_ties = FALSE)
}

# filter district level data to selected condition, grade level, and vaccination status
district_data_condition <- filter(district_data, disease_or_vaccine == condition,
                                immunization_status == vaccination_level,
                                grade == select_grade)



```

### Clean Names

School and district name formats differ between the vaccination data and the location data files, so we try to standardize the names to improve matching between the two files.

```{r}
# clean school names for better matching between vaccination file and location data
school_data_condition <- school_data_condition |> 
  # convert school name to lower case
  mutate(clean_name = str_to_lower(school_name),
         # remove text inside of parentheses
         clean_name = str_replace(clean_name, "\\(.*\\)", ""),
         # remove school as it is used inconsistently between the two data sources
         clean_name = str_replace(clean_name, 'school', ''),
         # replace elem with elementary
         clean_name = str_replace(clean_name, " elem", " elementary"),
         # replace & with and
         clean_name = str_replace(clean_name, "&", "and"),
         # remove |
         clean_name = str_replace(clean_name, "\\|", ""),
         # remove extra whitespace
         clean_name = str_squish(clean_name))

school_location <- school_location |> 
  # convert school name to lower case
  mutate(clean_name = str_to_lower(School),
         # remove text inside of parentheses
         clean_name = str_replace(clean_name, "\\(.*\\)", ""),
         # remove school as it is used inconsistently between the two data sources
         clean_name = str_replace(clean_name, 'school', ''),
         # replace elem with elementary
         clean_name = str_replace(clean_name, " elem", " elementary"),
         # replace prog with program
         clean_name = str_replace(clean_name, " prog$", " program"),
         # replace & with and
         clean_name = str_replace(clean_name, "&", "and"),
         # remove |
         clean_name = str_replace(clean_name, "\\|", ""),
         # remove extra whitespace
         clean_name = str_squish(clean_name))
        
district_data_condition <- district_data_condition |> 
  # clean up the name for one of the districts
  mutate(school_district = str_replace(school_district, "Hist\\.", "Historical"))

# combine school vaccination data and school location data using the created cleaned school name
combined_school <- left_join(school_data_condition, school_location, join_by(clean_name, school_district ==LEAName ))

# identify the number of schools in the vaccination data that could not be mapped to the school location file - counting total and just public schools
non_geocoded_schools_count <- nrow(filter(combined_school, is.na(SchoolCode)))
non_geocoded_pub_schools_count <- nrow(filter(combined_school, school_type == 'PUBLIC',
                                        is.na(SchoolCode)))
```

The shape data file only contains data on public schools and may not match all schools in the vaccination file. Including the private schools, there are still `r non_geocoded_schools_count` total schools in the vaccination data that we do not have location data for, `r non_geocoded_pub_schools_count` of which are public schools.

## Geocoding School Addresses

Geocoding converts a human-readable address (e.g. "123 Main St, Seattle, WA") into geographic coordinates (latitude and longitude) that can be plotted on a map. The `tidygeocoder` package provides a tidy interface to several geocoding services. We will use the [US Census geocoder](https://geocoding.geo.census.gov/geocoder/), which is free and requires no API key. **This should only be used on public data** as it is sending data to an outside service. We can use geocoding to identify the location of the schools that we only have addresses for and which we could not map to the geospatial location file.

### US Census Geocoder

```{r}

# I am only going to geocode a subset of the data for this example - so we will just geocode the public schools that we could not map to the location file
nongeocoded_public_schools <- filter(combined_school, is.na(SchoolCode),
                                     school_type == 'PUBLIC')

# add a state column for geocoding purposes
nongeocoded_public_schools <- nongeocoded_public_schools |> 
  mutate(state = 'Washington')

# pass the dataframe containing the address information (street, city, state, and zipcode) to tidygeocoder::geocode using the US Census API
# to geocode the address, which will return latitude/longitude data. The geocoder can accommodate abbreviations and some differences in format.
geocoded_public_schools <- nongeocoded_public_schools |>
  geocode(street = address,
          city = city,
          state = state,
          postalcode = zip,
          method = "census", # Using the US Census geocoding service - free, no API key required, can geocode up to 10,000 addresses
          lat = latitude,
          long = longitude)

# this may still be unable to geocode all the addresses - we will remove these for future plotting
geocoding_fail_count <- nrow(filter(geocoded_public_schools, is.na(latitude)))
geocoded_public_schools <- filter(geocoded_public_schools, !is.na(latitude))

# convert to a geometry shape dataframe
geocoded_public_schools <- st_as_sf(geocoded_public_schools, coords = c('longitude', 'latitude'), remove = FALSE)

```

**Note:** Some addresses may fail to geocode due to formatting issues, PO boxes, or addresses not found in the Census database. Inspect failures and consider cleaning those addresses manually or using a different geocoder. In our attempt `r geocoding_fail_count` addresses failed to geocode and were removed from our data.

### Combine Geocoding Data

```{r}
# combine all data, removing any data which were were unable to geocode
# combine the school vaccination data we geocoded using the shape file
combined_school_filtered <- filter(combined_school, !is.na(SchoolCode)) |> 
  rename(latitude = BldgLatitu,
         longitude = BldgLongit) |> 
  select(school_year, geography, grade, disease_or_vaccine, immunization_status,
         count, enrollment, percent, school_district, school_name, address, city, zip, school_type, county, SchoolCode,latitude, longitude, geometry)

# combine with the school vaccination data we geocoded using US Census geocoder
geocoded_school_filtered <-geocoded_public_schools |> 
         select(school_year, geography, grade, disease_or_vaccine, immunization_status,
                count, enrollment, percent, school_district, school_name, address, city, zip, school_type, county, SchoolCode,latitude, longitude, geometry)

combined_school_filtered <- rbind(combined_school_filtered, geocoded_school_filtered)

```

### Pull Washington Shape Data

The usmap package contains geographic data on all US counties, which can be used for creating a map plot.

```{r}
# pull US mapping data
us_county_map <- usmap::us_map('counties')
# subset to WA state data
WA_map_counties <- filter(us_county_map, abbr == 'WA') %>%
  # simplify county names
  mutate(county = str_replace(county, " County", ""))
# convert a geometry object to an sfc object for plotting purposes
WA_sfc <- st_as_sfc(WA_map_counties, crs = usmap_csv()@projargs)
# get subset of map object
# creates sf object, which extends data.frame like objects with a  simple feature list column
WA_sf_map <- st_sf(data.frame(fips = unique(WA_map_counties$fips), county = WA_map_counties$county, geometry = WA_sfc))
# find the centroid of each county
WA_sf_map$centroids <- st_centroid(WA_sf_map$geometry)

```

## Map Plot of WA School Vaccination Data

### All Washington Schools

```{r}

wa_vacc_map <- ggplot(WA_sf_map) +
  # use geom_sf to create map - pulls from geometry column in the data
  # set fill for each county
  geom_sf(fill = 'azure2', color = 'white') +
  # add county labels
  geom_sf_text(aes(label = county), size = 3, color = 'black') +
  # plot the school locations based on lat/long - size each point by the size of the school and color by the vaccination rate
  geom_point(data = filter(combined_school_filtered, !is.na(percent)), aes(longitude, latitude, fill = percent, size = enrollment),pch = 21, alpha = .8) +
  coord_sf(crs = st_crs(4283)) + #needed to make flat/horizontally aligned map for visualizing purposes
  scale_fill_viridis() +
  # set clean theme that removes axis labels
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5)) +
  # add title based on data filtered
  labs(fill = paste(condition, 'Vaccination Rate'), size = 'School Enrollment') +
  ggtitle(paste('Washington', condition, 'Vaccination Rates in Schools'))

wa_vacc_map
```

### Schools with Low Vaccination Rates

This plot above is difficult to visualize as there are many schools, which creates many overlapping data points. We can create a second plot that just shows schools with a vaccination rate less than 85%.

```{r}
wa_low_vacc_map <- ggplot(WA_sf_map) +
  # use geom_sf to create map - pulls from geometry column in the data
  # set fill for each county
  geom_sf(fill = 'azure2', color = 'white') +
  # add county labels
  geom_sf_text(aes(label = county), size = 3, color = 'black') +
  # filter to low vaccine rate schools and plot the school locations based on lat/long - size each point by the size of the school and color by the vaccination rate
  geom_point(data = filter(combined_school_filtered, !is.na(percent), percent < .85), aes(longitude, latitude, fill = percent, size = enrollment),pch = 21, alpha = .8) +
  coord_sf(crs = st_crs(4283)) + #needed to make flat/horizontally aligned map for visualizing purposes
  scale_fill_viridis() +
    # set clean theme that removes axis labels
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  # add title based on data filtered
  labs(fill = paste(condition, 'Vaccination Rate'), size = 'School Enrollment',
       title = paste('Washington', condition, 'Vaccination Rates in Schools'),
        subtitle = 'For Schools with a Vaccination Rate less than 85%')

wa_low_vacc_map
```

## Pulling GeoSpatial Data Using the API

### What is an API?

An **API** (Application Programming Interface) is a structured way to request data from a server over the internet. Instead of downloading a file manually, you send a request to a URL with parameters specifying what data you want, and the server returns data in the format specified (often JSON)

The Washington State Geospatial Open Data Portal exposes school district boundaries via an **ArcGIS REST API**. In the example above we downloaded the school location shape file from the Washington State Geospatial Open Data Portal and used it for plotting. We can also pull that data by querying the API, which ensures that our data is up to date.

### School District Boundaries

We will use the API to pull the school district boundary data from the [Washington State Geospatial Open Data Portal](https://geo.wa.gov/datasets/72ad21c67ecf4f21bc794d4d21485d86_0/explore?location=47.265238%2C-120.817601%2C7).

#### Metadata Query

```{r}
# URL for the school district data
service_URL <- "https://services9.arcgis.com/fWunDXKkvCx1CM4b/arcgis/rest/services/Washington_School_Districts/FeatureServer"

# we will pull the data, as well as info on when the data was published, the format, and record when we accessed the data
district_metadata <- list(service_URL = service_URL)

## Query metadata to get published date
# This builds and sends a GET request to the metadata endpoint of the API, and retrieve the response body as a plain string, and parse it as XML
metadata_req <- request(paste0(service_URL, "/info/metadata")) |>
  req_perform() |> # send the HTTP request
  resp_body_string() |> # extract the response as a raw string
  read_xml() # parse the string into XML

# extract the publication date from the metadata
esri_date <- metadata_req |>
  xml_find_first(".//Esri/CreaDate") |> # search XML for create data element
  xml_text() # extract that element as a string

# format the string returned into a data
district_metadata$publish_date <- paste(substr(esri_date, 1, 4),
                               substr(esri_date, 5, 6),
                               substr(esri_date, 7, 8),
                               sep = "-")

## Basic info used for our REST API queries
endpoint  <- paste0(service_URL, "/0/query")
crs <- 4326 # use the WGS84 coordinate system
response_format <- 'geojson' # format for returning the data
where_clause <- "1=1" # returns all records, no filtering

```

#### Fetch the Data

```{r}
## Retrieve the count from the server so you know how many records to expect
expected_count <- request(endpoint) |>
  req_url_query(where = where_clause, # filter to the records we set - in this case all records
                f = response_format, # select the respoinse format
                returnCountOnly = "true") |> # only return the count of records
  req_perform() |> # send the HTTP request and get the response
  resp_body_json() |> # parse th response as JSON into an R list
  _$properties$count # extract the count value from the list

cat(glue::glue("Expecting {expected_count} records"))

## Query the actual data - similarily filtering to all records and requesting geojson format data
district_shape_raw <- request(endpoint) |>
  req_url_query(where = where_clause,
                f = response_format,
                outFields = "*", # returns all attribute collumns
                outSR = crs) |> # set the output coordinate reference system
  req_perform() |> # send the HTTP request and get the response
  resp_body_string() # extract the response as a raw string

# Parse the raw geojson string into a nested R list, preserving the geometry structure
district_shape_dataset <- fromJSON(district_shape_raw, simplifyVector = FALSE)

## Check we got all expected records
received_count <- length(district_shape_dataset$features)
if(received_count != expected_count) {
  cat(glue::glue("Something went wrong with the request. Expected {expected_count} items, got {received_count}"))
} else {
   cat(glue::glue("Request succeeded. {received_count} records pulled"))
}
## print the fields from the data that was pulled
district_fields <- names(district_shape_dataset$features[[1]]$properties)
cat("Fields returned: ", paste(district_fields, collapse = ", "))

```

#### Converting to an `sf` Object

The API returns geometry as a list of polygon rings (sets of coordinates). We convert this into an `sf` object so we can use spatial operations and plotting.

```{r}
# Convert to sf object 
district_shape_sf <- read_sf(district_shape_raw)
message("Coordinate Reference System (CRS): ", st_crs(district_shape_sf)$input)

# store metadata
district_metadata$crs <- paste0("EPSG:", st_crs(district_shape_sf)$epsg)
district_metadata$accessed <- as.character(Sys.Date())

if (crs == 4326) {
  district_metadata$ggplot_crs <- "WGS84"    # equivalent of plate carrée / mercator note
} else {
  stop("This example was only designed for use with EPSG:4326")
}

# view the data we pulled
ggplot(district_shape_sf) +
  # use geom_sf to plot the shape data - will pull from the geometry field in the provided data
  geom_sf(fill = 'steelblue', color = 'white') +
  # set the coordinate system
  coord_sf(crs = st_crs(crs))  +
  # remove axis grid
  theme_void() +
  # center plot titles
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  # set labels
  labs(title = 'School District Boundary Data',
       subtitle = 'From WA Geospatial Portal')
```

## Visualizing School District Vaccination Data

### Merging Shape Data with District Vaccination Data

```{r}
# join the district boundary file and the district level vaccination data
district_condition_sf <- full_join(district_data_condition, district_shape_sf, join_by(school_district == LEAName_1))

# convert to sf for plotting
district_condition_sf <- st_sf(district_condition_sf)

```

### Plot School District Vaccination Data

```{r}

ggplot(district_condition_sf) +
  # use geom_sf to plot the shape data - will pull from the geometry field in the provided data
  # filling in the districts using the vaccination rate
  geom_sf(aes(fill = percent), color = 'white') +
  # set the coordinate system
  coord_sf(crs = st_crs(crs))  +
  # set a color palette for the fill
  scale_fill_viridis() +
  # remove axis lines
  theme_void() +
  # center and bold title
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5)) +
  # labels
  labs(title = paste('School District', condition, 'Vaccination Rates'),
       fill = 'Percent with\nComplete Vaccination',
       caption = 'gray districts are missing data')

```

### Combine School and District Data for a Single County

```{r}

selected_county <- "Mason"
# filter to selected county
ggplot(filter(district_condition_sf, County == selected_county)) +
  # use geom_sf to plot the shape data - will pull from the geometry field in the provided data
  # filling in the districts using the vaccination rate
  geom_sf(aes(fill = percent), color = 'white', alpha = .7) +
  # we could add district labels, although in practice they are hard to view
  # geom_sf_text(aes(label = school_district), size = 2, color = 'black') +
  # add in the schools for the selected county, filling by vaccination rate and sizing by enrollment
  geom_point(data = filter(combined_school_filtered, !is.na(percent), county == selected_county),
             aes(longitude, latitude, fill = percent, size = enrollment),
             pch = 21, alpha = .8) +
  # set coordinate system
  coord_sf(crs = st_crs(crs)) +
  # fill color palette
  scale_fill_viridis() +
  # remove axis lines
  theme_void() +
  # center and format titles
  theme(plot.title = element_text(size = 15, face = "bold",hjust = 0.5),
         plot.subtitle = element_text(size = 13, hjust = 0.5)) +
  # labels
  labs(title = paste('School and School District', condition, 'Vaccination Rates'),
       subtitle = paste(selected_county, 'County'),
       fill = 'Percent with\nComplete Vaccination')
```
